Spatial Reasoning and Statistical Graphics

### 2014-10-29 Paper/Draft1/TestingVisualAptitude.Rnw
starting to be in a readable format again

options(replace.assign=TRUE,width=70,scipen=3)
require(knitr)

library(reshape2)
suppressMessages(library(ggplot2))
library(plyr)
suppressMessages(library(gridExtra))
library(xtable)
source(paste(codedir, "Analysis.R", sep=""))

@@

\comment{Intro about statistical graphics, lineups, ...}
\subsection{Spatial Reasoning and Statistical Graphics}
Statistical graphics provide quick summaries of data, models, and results, but these displays may not be equally useful to all viewers. Mathematical ability is important, even for the simplest graphs: XXX in \cite{shah1995conceptual} showed that mathematical ability, but not spatial ability, was associated with accuracy on a simple two-dimensional line graph. Spatial ability becomes more important, however, when more complicated graphical displays are used in comparison tasks: XXX in \cite{mayer1994whom} hypothesize that individuals with low spatial ability perform worse on tests utilizing diagrams and graphs because more cognitive resources are required to process the visual stimuli, which leaves fewer resources to make connections and draw conclusions from those stimuli. Many theories of graphical learning center around the difference between visual and verbal processing: the dual-coding theory emphasizes the utility of complementary information in both domains, while the visual argument hypothesis emphasizes that graphics are more efficient tools for providing data with spatial, temporal, or other implicit ordering, because the spatial dimension can be represented graphically in a more natural manner \cite{vekiri2002value}. Both of these theories suggest spatial ability would impact a viewer's use of graphics, because spatial ability either influences cognitive resource allocation or affects the processing of spatial relationships between graphical elements. Previous investigations into graphical learning and spatial ability have found relationships between spatial ability and the ability to read information from graphs \cite{lowrie2007solving}.

@@

Statistical lineups \cite{hofmann2012graphical,buja2009statistical} depend on the ability to search for a signal amid distractors (visual search) and the ability to infer patterns from stimuli (pattern recognition). Some lineups (polar coords) also depend on the ability to mentally rotate stimuli (spatial rotation) and mentally manipulate graphs (spatial rotation and manipulation). By breaking the lineup task down into component parts, we may be able to determine which visuospatial factors most strongly correlate with lineup performance, using carefully chosen cognitive tests to assess these aspects of visuospatial ability. These assessments are then used to account for some of the variation in lineup performance differences. In addition, we also examine the effect that previous experiences (science-based major, research experience, Auto-CAD skills) have on lineup performance and the implications for the use of lineups to assess the graphical display of statistical information.

@@

The \textbf{visual searching task} (VST), shown in figure~\ref{fig:VST}, is designed to test participants' ability to find a target stimulus in a field of distractors. The visual search task is similar in concept to lineups: it tests an individual's ability to find the target plot. Historically, it has been used as a measure of brain damage\cite{goldstein1973validity,demita1981validity,moerland1986neuropsychological}; however, similar tasks, have been used to measure cognitive performance in a variety of situations (under the influence of drugs, for example, in XXX in \cite{anderson1983interactive}). The similarity to the lineup protocol as well as the simplicity of the test and its' lack of color justify the slight deviation from forms of visual search tasks typically used in normal populations. 
\begin{figure}[hbpt]\centering
\includegraphics[width=\linewidth]{VisualSearch}
\caption[Visual Search Task]{Visual Search Task. Participants are instructed to find the plot numbered 1-24 which matches the plot labeled ``Target". Participants will complete up to 25 of these tasks in 5 minutes.}\label{fig:VST}
\end{figure}

The \textbf{figure~classification task} tests a participant's ability to extrapolate rules from provided figures. This task is associated with inductive reasoning ablities (factor I in XXX in \cite{ekstrom1976manual}). An example is shown in figure~\ref{fig:figureclassification}. 
\begin{figure*}[hbpt]\centering
\includegraphics[width=.75\linewidth]{figureclassification}
\caption[Figure~Classification Task]{Figure~Classification Task. Participants classify each figure~in the second row as belonging to group 1, 2, or 3 (if applicable). Participants will complete up to 14 of these tasks (each consisting of 8 figures to classify) in 8 minutes.}\label{fig:figureclassification}
\end{figure*}
The figure~classification test requires the same type of reasoning as the lineups: participants must determine the rules from the provided classes, and extrapolate from those rules to classify new figures. In lineups, participants must determine the rules based on the panels appearing in the lineup; they must then identify the plot which does not conform. As such, the figure~classification test has content validity in relation to lineup performance: it is measuring similar underlying criteria.

The \textbf{card rotation test} measures a participant's ability to rotate objects in two dimensions in order to distinguish between left-hand and right-hand versions of the same figure. It tests mental rotation skills, and is classified as a test of spatial orientation in XXX in \cite{ekstrom1976manual}, though it does require that participants have both mental rotation ability and short-term visual memory. An example is shown in figure~\ref{fig:cardrotation}.
\begin{figure*}[hbpt]\centering
\includegraphics[width=.75\linewidth]{cardrotation}
\caption[Card Rotation Task]{Card Rotation Task. Participants mark each figure~on the right hand side as either the same or different than the figure~on the left hand side of the dividing line. Participants will complete up to 20 of these tasks (each consisting of 8 figures) in 6 minutes.}\label{fig:cardrotation}
\end{figure*}
Two-dimensional comparisons are an important component of lineup performance. In some lineup situations, these comparisons involve translation, but in other lineups, rotation is required. Lineups also require visual short-term memory, so the additional factor measured implicitly by this test does not reduce its potential relevance to lineup performance.

The \textbf{paper folding test} measures participants' ability to visualize and mentally manipulate figures in three dimensions. It is classified as part of the visualization factor in XXX in \cite{ekstrom1976manual}, which differs from the spatial orientation factor because it requires participants to visualize, manipulate, and transform the figure~mentally, which makes it a more complex and demanding task than simple rotation. The paper folding test is associated with the ability to extrapolate symmetry and reflection over multiple steps. An example is shown in figure~\ref{fig:paperfolding}. 
\begin{figure*}[hbpt]\centering
\includegraphics[width=.75\linewidth]{paperfolding}
\caption[Paper Folding Task]{Paper Folding Task. Participants are instructed to pick the figure~matching the sequence of steps shown in the left-hand figure. Participants will complete up to 20 of these tasks in 6 minutes.}\label{fig:paperfolding}
\end{figure*}
Lineups require similar manipulations in two-dimensional space, and also require the ability to perform complex spatial manipulations mentally; for instance, comparing the interquartile range of two boxplots as well as their relative alignment to a similar set of two boxplots in another panel.

Between cognitive tasks, participants will also complete three blocks of 20 lineups each, assembled from previous studies \cite{hofmann2012graphical,majumder2013validation}. Participants have 5 minutes to complete each block of 20 lineups. Figure~\ref{fig:lineup} shows a sample lineup of box plots.

@@

Scoring of all test results was done such that random guessing leads to an expected value of 0; therefore each question answered correctly contributes to the score by 1, while a wrong answer is scored by $-1/(k-1)$, where $k$ is the total number of possible answers to the question. Thus, for a test consisting of multiple choice questions with $k$ suggested answers with a single correct answer each, the score is calculated as
\begin{eqnarray}\label{eq.scoring}
\# \text{correct answers } - 1/(k-1) \cdot \# \text{wrong answers}.
\end{eqnarray}
This allows us to compare each participant's score in light of how many problems were attempted as well as the number of correct responses. Combining accuracy and speed into a single number does not only make a comparison of test scores easier,  this scoring mechanism is also used on many standardized tests, such as the SAT and the battery of psychological tests \cite{diamond1973correction, ekstrom1976manual} from which parts of this test are drawn. The advantage of using tests from the Kit of Factor Referenced Cognitive tests \cite{ekstrom1976manual} is that the tests are extremely well studied (including an extensive meta-analysis by XXX in \cite{voyer1995magnitude} of the spatial tests we are using in this study) and comparison data are available from the validation of these factors \cite{schaie1998longitudinal,hampson1990variations,mayer1994whom} and previous versions of the kit \cite{educational1963kit}.
\comment{maybe reference book by Michel Hersen\\\textit{I was working on it yesterday along with trying to validate the general idea that reading graphics requires certain spatial abilities. It's definitely a work in progress. I haven't seen that book before, but I'm not at all surprised it exists.}}

@@

\subsection{Comparison of Spatial Tests with Previously Validated Results}
The card rotation, paper folding, and figure~classification tests have been validated using different populations, many of which are demographically similar to Iowa State students (naval recruits, college students, late high-school students, and 9th grade students). We compare Iowa State students' unscaled scores in table \ref{tab:scorecomparison}, adjusting data from other populations to account for subpopulation structure and test length.

\begin{savenotes}
\begin{table*}[ht]
\centering
\begin{tabular}{rlllc}
\hline
  & Card Rotation & Paper Folding & Figure~Classification & Visual Search  \\\hline
ISU Students & \Sexpr{round(mean(ans.summary$card_rot), 1)} (\Sexpr{round(sd(ans.summary$card_rot), 1)}) 
             & \Sexpr{round(mean(ans.summary$folding), 1)} (\Sexpr{round(sd(ans.summary$folding), 1)})
             & \Sexpr{sprintf("%.1f", mean(ans.summary$fig_class))} (\Sexpr{round(sd(ans.summary$fig_class), 1)})\footnote{ISU students took only Part I due to time constraints.}
             & \Sexpr{round(mean(ans.summary$vis_search), 1)} (\Sexpr{round(sd(ans.summary$vis_search), 1)})\\
Scaled Scores & 88.0 (\Sexpr{round(sqrt((24.6^2)*2), 1)})
              & 13.8 (4.5)
              & \Sexpr{round((120.0*294 + 114.9*323)/617/2, 1)} (\Sexpr{round(sqrt((30^2*294 + 27.8^2*323)/617)/2, 1)})\footnote{Averages calculated assuming 294 males and 323 females.}
              & -- \\
Unscaled Scores & 44.0 (24.6)\footnote{Data from Part I only.}
                & 13.8 (4.5)
                & M:~120.0~(30.0), F:~114.9~(27.8)
                & --\\
{\footnotesize Population}    
              & \parbox[t]{.15\linewidth}{\footnotesize approx.\ 550 male\\ naval recruits} 
              & \parbox[t]{.15\linewidth}{\footnotesize 46 college students\\(1963~version)}
              & \parbox[t]{.3\linewidth}{\footnotesize suburban 11th \& 12th grade students\\(288-300 males, 317-329 females)}
              & \\\hline
\end{tabular}

\caption{Comparison of scores from Iowa State students and scores reported in XXX in \cite{ekstrom1976manual}. Scaled scores are calculated based on information reported in the manual, scaled to account for differences in the number of questions answered during this experiment. Data shown are from the population most similar to ISU students, out of the data available. The Visual Search task \cite{goldstein1973validity,demita1981validity,moerland1986neuropsychological} is not part of the Kit of Factor Referenced Cognitive Test data, and thus we do not have comparison data for the form used in this experiment. \label{tab:scorecomparison}}
\end{table*}
\end{savenotes}

Table \ref{tab:scorecomparison} shows mean scores and standard deviation for ISU students and other populations. Values have been adjusted to accommodate for differences in test procedures and sub-population structure; for instance,  some data is reported for a single part of a two-part test, or results are reported for each gender separately (adjustment procedure is described in more detail in Appendix \ref{app:ScoreAdj}). Once these adjustments have been completed, it is evident that Iowa State undergraduates scored at about the same level as other similar demographics. In fact, both means and standard deviations of ISU students' scores are similar to the comparison groups, which were chosen from available demographic groups based on population similarity.

@@

Applying the grading protocol discussed in section \ref{sec:scaling}, we see that the ranges of lineup and visuospatial test scores do not include zero; this indicates that we do not see random guessing from participants in any task. Figure~\ref{fig:Scores} shows the range of possible scores and the observed score distribution.

<<ResultsScaledScores, echo=F, include=T, fig.pos="hbt", fig.cap="Test scores for lineups and visuospatial tests. As none of the participants scored at or below zero, we can conclude that there is little evidence of random guessing. We also note the score compression that occurs on the Visual Search test; this indicates that most participants scored extremely high, and thus, participants' scores are not entirely representative of their ability. \\label{fig:Scores}", fig.width=8, fig.height=3, out.width='.95\\linewidth', warning=F, message=F>>=

score.summary <- melt(ans.summary, measure.vars=20:24, value.name = "Score", variable.name = "Test")
score.summary$Test <- factor(score.summary$Test, levels=c("lineup", "card_rot", "fig_class", "folding", "vis_search"), labels=c("Lineups", "Card Rotation", "Figure~Class.", "Paper Folding", "Visual Search"))
qrange$sqrt.k <- sqrt(qrange$k-1)
qrange$Test <- factor(qrange$testtype, levels=c("lineup", "card_rot", "fig_class", "folding", "vis_search"), labels=c("Lineups", "Card Rotation", "Figure~Class.", "Paper Folding", "Visual Search"))
ggplot() + 
  geom_bar(data=qrange, aes(width=.9, x=Test, y=-n/(k-1), fill="red"), alpha=.5, colour="red", stat="identity", inherit.aes=F) + 
  geom_bar(data=qrange, aes(width=.9, x=Test, y=n, colour="red"), stat="identity", fill=NA, inherit.aes=F) + 
  geom_boxplot(data=score.summary, aes(x=factor(Test), y=Score), inherit.aes=F, fill="grey") +
  scale_fill_discrete(name="", labels="Scores indicating guessing") +
  scale_colour_discrete(name="", labels="Theoretical Range") + 
  theme_bw()  + 
  theme(legend.position="bottom", legend.box="horizontal") + 
  ylab("Test Scores") + 
  xlab("") +
  coord_flip()
@
%

\comment{

\begin{enumerate}
\item some general words on behavior of lineup scores by themselves - I think the whole discussion that is at the moment  at the end of the paper should move up here. The levels of the video gaming are a bit messed up - they should be ordered according to number of games played rather than alphabetically :) - and I think that then the relationship actually is a monotonic one ...
 In any case, sort the figures in the table according to their $p$-values in the table. Gender shows up as significantly different in my table of $p$ values, but in the text you say that it is not. 
 The table with the t-tests should probably go into an appendix .... 
\item figure~8 with either three or four scatterplots depending on how the visual search scores fit in
\item a paragraph on discussing the multicollinearity between the test scores based on correlations (either, again, including visual search or not) - this should lead to the conclusion that we cannot directly use a linear model of lineup in test scores.
\item PCA of test scores, with a bit of discussion
\item linear model of lineup scores in PCA with discussion and tying it back to raw scores (figure~classification really is the best predictor for lineups, but it only explains some of the variance).
\end{enumerate}
 }

@@

<<VisReasoningCategorical, echo=F, include=F, eval=T, fig.width=8,fig.height=5, message=F, warning=F>>=
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="verbal_skills", value=1))
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="math_skills", value=1))
lineup.summary.categorical$variable <- factor(lineup.summary.categorical$variable, labels=c("Age", "Sex", "Math/Science Research", "Statistics Class", "Calculus 1", "Verbal Skills", "Math Skills", "Art Skills", "AutoCAD Experience", "Video Game Hrs/Wk"))
lineup.summary.categorical$value[lineup.summary.categorical$value=="f"] <- "Female"
lineup.summary.categorical$value[lineup.summary.categorical$value=="m"] <- "Male"
lineup.summary.categorical$value[lineup.summary.categorical$value=="y"] <- "TRUE"
lineup.summary.categorical$value[lineup.summary.categorical$value=="n"] <- "FALSE"

@@

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="boxplot") + facet_wrap(~variable, scales="free_x", ncol=5) + xlab("") + ylab("Scaled Lineup Score") + theme_bw() + geom_point(aes(x=value, y=lineup), shape=1, size=3)
@
\begin{figure*}[hbpt]\centering
\includegraphics[width=\linewidth]{fig-VisReasoningCategorical-1}
\caption[Visual Aptitude Study Results]{Sample demographic characteristics compared with lineup score. Categories are ordered by effect size; calculus completion,  hours spent playing video games per week, and sex are all associated with a difference in lineup score. }\label{fig:visualaptitudecat}
\end{figure*}

Completion of Calculus I is associated with increased performance on lineups. This may be related to general math education level, or it may be that success in both lineups and calculus requires certain visual skills. This association is consistent with \cite{shah1995conceptual}, who found an association between  mathematical ability and performance on simple graph description tasks.  There is also a significant association between hours of video games played per week and score on lineups, however, this association is not monotonic and the groups do not have equal sample size, so the conclusion may be suspect. There is a (nearly) significant difference between male and female performance on lineups; this is not particularly surprising, since men perform better on many spatial tests \cite{voyer1995magnitude} and performance on spatial tests is correlated with phase of the menstrual cycle in women \cite{hausmann2000sex}. There is no significant difference in lineup performance for participants of different age, self-assessed skills in various domains, previous participation in math or science research, completion of a statistics class, or experience with AutoCAD. These demographic characteristics were chosen to account for life experience and personal skills which may have influenced the results. Statistical test results are available in appendix \ref{app:categoricalresults}.

@@

\item XXX in \cite{just1985cognitive} showed that high-spatial-ability viewers used different rotation strategies than low-spatial-ability viewers when asked to whether three-dimensional alphabet cubes were the same. 
\item XXX in \cite{voyer1995magnitude} completed a meta-analysis of spatial tasks and gender differences. They concluded that ``rotation" tasks, such as the card rotation test, have robust sex differences across cohort, and while the sex differences seem to be declining for some tests, such as the card rotation test, the differences still exist for the time being. Test administration differences may account for the change. They also concluded that the paper-folding test showed no such sex difference, though this may be dependent on test scoring (a lower guessing penalty decreases the score difference). 

@@

\item XXX in \cite{lowrie2007solving} investigated students (9-10 yrs) ability to read mathematical graphs and their visual-spatial ability. They partition graphs into 6 ``graphical languages" (some of these languages aren't statistical graphs - networks, venn diagrams, maps), and use multivariate analysis to associate spatial ability with some of the types of graphs studied. 

@@

They also cite a bunch of studies (including XXX in \cite{voyer1995magnitude}) that I haven't had time to completely explore.}}

>>>>

options(replace.assign=TRUE,width=70,scipen=3)

library(reshape2)
suppressMessages(library(ggplot2))
library(plyr)
library(grid)
suppressMessages(library(gridExtra))
library(xtable)
source(paste(codedir, "Analysis.R", sep=""))

@@

\comment{Intro about statistical graphics, lineups, ...}
\subsection{Spatial Reasoning and Statistical Graphics}
Statistical graphics provide quick summaries of data, models, and results, but these displays may not be equally useful to all viewers. Mathematical ability is important, even for the simplest graphs:  mathematical ability, not spatial ability, was shown~\cite{shah1995conceptual} to be associated with accuracy on a simple two-dimensional line graph. Spatial ability becomes more important, however, when more complicated graphical displays are used in comparison tasks: the lower performance of  individuals with low spatial ability  on tests utilizing diagrams and graphs is attributed ~\cite{mayer1994whom} to the fact that more cognitive resources are required to process the visual stimuli, which leaves fewer resources to make connections and draw conclusions from those stimuli. Many theories of graphical learning center around the difference between visual and verbal processing: the dual-coding theory emphasizes the utility of complementary information in both domains, while the visual argument hypothesis emphasizes that graphics are more efficient tools for providing data with spatial, temporal, or other implicit ordering, because the spatial dimension can be represented graphically in a more natural manner~\cite{vekiri2002value}. Both of these theories suggest spatial ability would impact a viewer's use of graphics, because spatial ability either influences cognitive resource allocation or affects the processing of spatial relationships between graphical elements. Previous investigations into graphical learning and spatial ability have found relationships between spatial ability and the ability to read information from graphs~\cite{lowrie2007solving}.

@@

Statistical lineups~\cite{hofmann2012graphical,buja2009statistical} depend on the ability to search for a signal amid distractors (visual search) and the ability to infer patterns from stimuli (pattern recognition). Some lineups (polar coords) also depend on the ability to mentally rotate stimuli (spatial rotation) and mentally manipulate graphs (spatial rotation and manipulation). By breaking the lineup task down into component parts, we may be able to determine which visuospatial factors most strongly correlate with lineup performance, using carefully chosen cognitive tests to assess these aspects of visuospatial ability. These assessments are then used to account for some of the variation in lineup performance differences. In addition, we also examine the effect that previous experiences (science-based major, research experience, Auto-CAD skills) have on lineup performance and the implications for the use of lineups to assess the graphical display of statistical information.

@@

The \textbf{visual searching task} (VST), shown in figure~\ref{fig:VST}, is designed to test participants' ability to find a target stimulus in a field of distractors. The visual search task is similar in concept to lineups: it tests an individual's ability to find the target plot. Historically, it has been used as a measure of brain damage\cite{goldstein1973validity,demita1981validity,moerland1986neuropsychological}; however, similar tasks, have been used to measure cognitive performance in a variety of situations (under the influence of drugs, for example, in~\cite{anderson1983interactive}). The similarity to the lineup protocol as well as the simplicity of the test and its' lack of color justify the slight deviation from forms of visual search tasks typically used in normal populations. 
\begin{figure}[hbpt]\centering
\includegraphics[width=\linewidth]{VisualSearch}
\caption[Visual Search Task]{Visual Search Task. Participants are instructed to find the plot numbered 1-24 which matches the plot labeled ``Target". Participants will complete up to 25 of these tasks in 5 minutes.}\label{fig:VST}
\end{figure}

The \textbf{figure~classification task} tests a participant's ability to extrapolate rules from provided figures. This task is associated with inductive reasoning ablities (factor I  in~\cite{ekstrom1976manual}). An example is shown in figure~\ref{fig:figureclassification}. 
\begin{figure*}[hbpt]\centering
\includegraphics[width=.75\linewidth]{figureclassification}
\caption[Figure~Classification Task]{Figure~Classification Task. Participants classify each figure~in the second row as belonging to group 1, 2, or 3 (if applicable). Participants will complete up to 14 of these tasks (each consisting of 8 figures to classify) in 8 minutes.}\label{fig:figureclassification}
\end{figure*}
The figure~classification test requires the same type of reasoning as the lineups: participants must determine the rules from the provided classes, and extrapolate from those rules to classify new figures. In lineups, participants must determine the rules based on the panels appearing in the lineup; they must then identify the plot which does not conform. As such, the figure~classification test has content validity in relation to lineup performance: it is measuring similar underlying criteria.

The \textbf{card rotation test} measures a participant's ability to rotate objects in two dimensions in order to distinguish between left-hand and right-hand versions of the same figure. It tests mental rotation skills, and is classified as a test of spatial orientation in~\cite{ekstrom1976manual}, though it does require that participants have both mental rotation ability and short-term visual memory. An example is shown in figure~\ref{fig:cardrotation}.
\begin{figure*}[hbpt]\centering
\includegraphics[width=.75\linewidth]{cardrotation}
\caption[Card Rotation Task]{Card Rotation Task. Participants mark each figure~on the right hand side as either the same or different than the figure~on the left hand side of the dividing line. Participants will complete up to 20 of these tasks (each consisting of 8 figures) in 6 minutes.}\label{fig:cardrotation}
\end{figure*}
Two-dimensional comparisons are an important component of lineup performance. In some lineup situations, these comparisons involve translation, but in other lineups, rotation is required. Lineups also require visual short-term memory, so the additional factor measured implicitly by this test does not reduce its potential relevance to lineup performance.

The \textbf{paper folding test} measures participants' ability to visualize and mentally manipulate figures in three dimensions. It is classified as part of the visualization factor in~\cite{ekstrom1976manual}, which differs from the spatial orientation factor because it requires participants to visualize, manipulate, and transform the figure~mentally, which makes it a more complex and demanding task than simple rotation. The paper folding test is associated with the ability to extrapolate symmetry and reflection over multiple steps. An example is shown in figure~\ref{fig:paperfolding}. 
\begin{figure*}[hbpt]\centering
\includegraphics[width=.75\linewidth]{paperfolding}
\caption[Paper Folding Task]{Paper Folding Task. Participants are instructed to pick the figure~matching the sequence of steps shown in the left-hand figure. Participants will complete up to 20 of these tasks in 6 minutes.}\label{fig:paperfolding}
\end{figure*}
Lineups require similar manipulations in two-dimensional space, and also require the ability to perform complex spatial manipulations mentally; for instance, comparing the interquartile range of two boxplots as well as their relative alignment to a similar set of two boxplots in another panel.

Between cognitive tasks, participants will also complete three blocks of 20 lineups each, assembled from previous studies~\cite{hofmann2012graphical,majumder2013validation}. Participants have 5 minutes to complete each block of 20 lineups. Figure~\ref{fig:lineup} shows a sample lineup of box plots.

@@

Scoring of all test results was done such that random guessing leads to an expected value of 0; therefore each question answered correctly contributes to the score by 1, while a wrong answer is scored by $-1/(k-1)$, where $k$ is the total number of possible answers to the question. Thus, for a test consisting of multiple choice questions with $k$ suggested answers with a single correct answer each, the score is calculated as
\begin{eqnarray}\label{eq.scoring}
\# \text{correct answers } - 1/(k-1) \cdot \# \text{wrong answers}.
\end{eqnarray}
This allows us to compare each participant's score in light of how many problems were attempted as well as the number of correct responses. Combining accuracy and speed into a single number does not only make a comparison of test scores easier,  this scoring mechanism is also used on many standardized tests, such as the SAT and the battery of psychological tests~\cite{diamond1973correction, ekstrom1976manual} from which parts of this test are drawn. The advantage of using tests from the Kit of Factor Referenced Cognitive tests~\cite{ekstrom1976manual} is that the tests are extremely well studied (including an extensive meta-analysis in~\cite{voyer1995magnitude} of the spatial tests we are using in this study) and comparison data are available from the validation of these factors~\cite{schaie1998longitudinal,hampson1990variations,mayer1994whom} and previous versions of the kit~\cite{educational1963kit}.
\comment{maybe reference book by Michel Hersen\\\textit{I was working on it yesterday along with trying to validate the general idea that reading graphics requires certain spatial abilities. It's definitely a work in progress. I haven't seen that book before, but I'm not at all surprised it exists.}}

@@

\subsection{Comparison of Spatial Tests with Previously Validated Results}
The card rotation, paper folding, and figure~classification tests have been validated using different populations, many of which are demographically similar to Iowa State students (naval recruits, college students, late high-school students, and 9th grade students). We compare Iowa State students' unscaled scores in table~\ref{tab:scorecomparison}, adjusting data from other populations to account for subpopulation structure and test length.

\begin{savenotes}
\begin{table*}[ht]
\caption{Comparison of scores from Iowa State students and scores reported in~\cite{ekstrom1976manual}. Scaled scores are calculated based on information reported in the manual, scaled to account for differences in the number of questions answered during this experiment. Data shown are from the population most similar to ISU students, out of the data available. The Visual Search task~\cite{goldstein1973validity,demita1981validity,moerland1986neuropsychological} is not part of the Kit of Factor Referenced Cognitive Test data, and thus we do not have comparison data for the form used in this experiment. \label{tab:scorecomparison}}
\centering
\begin{tabular}{rlllc}
\hline
  & Card Rotation & Paper Folding & Figure~Classification & Visual Search  \\\hline
ISU Students & \Sexpr{round(mean(ans.summary$card_rot), 1)} (\Sexpr{round(sd(ans.summary$card_rot), 1)}) 
             & \Sexpr{round(mean(ans.summary$folding), 1)} (\Sexpr{round(sd(ans.summary$folding), 1)})
             & \Sexpr{sprintf("%.1f", mean(ans.summary$fig_class))} (\Sexpr{round(sd(ans.summary$fig_class), 1)})\footnote{ISU students took only Part I due to time constraints.}
             & \Sexpr{round(mean(ans.summary$vis_search), 1)} (\Sexpr{round(sd(ans.summary$vis_search), 1)})\\
Scaled Scores & 88.0 (\Sexpr{round(sqrt((24.6^2)*2), 1)})
              & 13.8 (4.5)
              & \Sexpr{round((120.0*294 + 114.9*323)/617/2, 1)} (\Sexpr{round(sqrt((30^2*294 + 27.8^2*323)/617)/2, 1)})\footnote{Averages calculated assuming 294 males and 323 females.}
              & -- \\
Unscaled Scores & 44.0 (24.6)\footnote{Data from Part I only.}
                & 13.8 (4.5)
                & M:~120.0~(30.0), F:~114.9~(27.8)
                & --\\
{\footnotesize Population}    
              & \parbox[t]{.15\linewidth}{\footnotesize approx.\ 550 male\\ naval recruits} 
              & \parbox[t]{.15\linewidth}{\footnotesize 46 college students\\(1963~version)}
              & \parbox[t]{.3\linewidth}{\footnotesize suburban 11th \& 12th grade students\\(288-300 males, 317-329 females)}
              & \\\hline
\end{tabular}                
\end{table*}
\end{savenotes}

Table~\ref{tab:scorecomparison} shows mean scores and standard deviation for ISU students and other populations. Values have been adjusted to accommodate for differences in test procedures and sub-population structure; for instance,  some data is reported for a single part of a two-part test, or results are reported for each gender separately (adjustment procedure is described in more detail in Appendix~\ref{app:ScoreAdj}). Once these adjustments have been completed, it is evident that Iowa State undergraduates scored at about the same level as other similar demographics. In fact, both means and standard deviations of ISU students' scores are similar to the comparison groups, which were chosen from available demographic groups based on population similarity.

@@

Applying the grading protocol discussed in section~\ref{sec:scaling}, we see that the ranges of lineup and visuospatial test scores do not include zero; this indicates that we do not see random guessing from participants in any task. Figure~\ref{fig:Scores} shows the range of possible scores and the observed score distribution.

<<ResultsScaledScores, echo=F, include=T, fig.pos="hbt", fig.cap="Test scores for lineups and visuospatial tests. As none of the participants scored at or below zero, we can conclude that there is little evidence of random guessing. We also note the score compression that occurs on the Visual Search test; this indicates that most participants scored extremely high, and thus, participants' scores are not entirely representative of their ability. \\label{fig:Scores}", fig.width=4.5, fig.height=4.5, out.width='\\linewidth', warning=F, message=F>>=

score.summary <- melt(ans.summary, measure.vars=20:24, value.name = "Score", variable.name = "Test")
score.summary$Test <- factor(score.summary$Test, levels=c("lineup", "card_rot", "fig_class", "folding", "vis_search"), labels=c("Lineups", "Card Rotation", "Figure~Class.", "Paper Folding", "Visual Search"))
qrange$sqrt.k <- sqrt(qrange$k-1)
qrange$Test <- factor(qrange$testtype, levels=c("lineup", "card_rot", "fig_class", "folding", "vis_search"), labels=c("Lineups", "Card Rotation", "Figure~Class.", "Paper Folding", "Visual Search"))
ggplot() + 
  geom_bar(data=qrange, aes(width=.9, x=Test, y=-n/(k-1), fill="red"), alpha=.5, colour="red", stat="identity", inherit.aes=F) + 
  geom_bar(data=qrange, aes(width=.9, x=Test, y=n, colour="red"), stat="identity", fill=NA, inherit.aes=F) + 
  geom_boxplot(data=score.summary, aes(x=factor(Test), y=Score), inherit.aes=F, fill="grey") +
  scale_colour_discrete(name="", labels="Theoretical\nRange") +
  scale_fill_discrete(name="", labels="Scores\nindicating\nguessing") +
  theme_bw()  + 
#  theme(legend.position="bottom", legend.box="horizontal") + 
  ylab("Test Scores") + 
  xlab("") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(plot.margin=unit(c(0,0,0,0),"mm"))
  #coord_flip()
@
%

@@

<<VisReasoningCategorical, echo=F, include=F, eval=T, fig.width=9,fig.height=5, message=F, warning=F>>=
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="verbal_skills", value=1))
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="math_skills", value=1))
lineup.summary.categorical$variable <- factor(lineup.summary.categorical$variable, labels=c("Age", "Sex", "Math/Science Research", "Statistics Class", "Calculus 1", "Verbal Skills", "Math Skills", "Art Skills", "AutoCAD Experience", "Video Game Hrs/Wk"))
lineup.summary.categorical$value[lineup.summary.categorical$value=="f"] <- "Female"
lineup.summary.categorical$value[lineup.summary.categorical$value=="m"] <- "Male"
lineup.summary.categorical$value[lineup.summary.categorical$value=="y"] <- "TRUE"
lineup.summary.categorical$value[lineup.summary.categorical$value=="n"] <- "FALSE"

@@

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="boxplot") + 
  facet_wrap(~variable, scales="free_x", ncol=5) + xlab("") + 
  ylab("Scaled Lineup Score") + 
  geom_point(aes(x=value, y=lineup), shape=1, size=3) + 
  theme_bw() + 
  theme(plot.margins=unit(c(0,0,0,0), unit="mm"))
@
\begin{figure*}[hbpt]\centering
\includegraphics[width=.9\linewidth]{fig-VisReasoningCategorical-1}
\caption[Visual Aptitude Study Results]{Demographic characteristics of participants compared with lineup score. Categories are ordered by effect size; calculus completion,  hours spent playing video games per week, and sex are all associated with a difference in lineup score. }\label{fig:visualaptitudecat}
\end{figure*}

Completion of Calculus I is associated with increased performance on lineups. This may be related to general math education level, or it may be that success in both lineups and calculus requires certain visual skills. This association is consistent with~\cite{shah1995conceptual}, who found an association between  mathematical ability and performance on simple graph description tasks.  There is also a significant association between hours of video games played per week and score on lineups, however, this association is not monotonic and the groups do not have equal sample size, so the conclusion may be suspect. There is a (nearly) significant difference between male and female performance on lineups; this is not particularly surprising, since men perform better on many spatial tests~\cite{voyer1995magnitude} and performance on spatial tests is correlated with phase of the menstrual cycle in women~\cite{hausmann2000sex}. There is no significant difference in lineup performance for participants of different age, self-assessed skills in various domains, previous participation in math or science research, completion of a statistics class, or experience with AutoCAD. These demographic characteristics were chosen to account for life experience and personal skills which may have influenced the results. Statistical test results are available in appendix~\ref{app:categoricalresults}.

@@

\item authors in~\cite{just1985cognitive} showed that high-spatial-ability viewers used different rotation strategies than low-spatial-ability viewers when asked to whether three-dimensional alphabet cubes were the same. 
\item authors in~\cite{voyer1995magnitude} completed a meta-analysis of spatial tasks and gender differences. They concluded that ``rotation" tasks, such as the card rotation test, have robust sex differences across cohort, and while the sex differences seem to be declining for some tests, such as the card rotation test, the differences still exist for the time being. Test administration differences may account for the change. They also concluded that the paper-folding test showed no such sex difference, though this may be dependent on test scoring (a lower guessing penalty decreases the score difference). 

@@

\item authors in~\cite{lowrie2007solving} investigated students (9-10 yrs) ability to read mathematical graphs and their visual-spatial ability. They partition graphs into 6 ``graphical languages" (some of these languages aren't statistical graphs - networks, venn diagrams, maps), and use multivariate analysis to associate spatial ability with some of the types of graphs studied. 

@@

They also cite a bunch of studies (including authors in~\cite{voyer1995magnitude}) that I haven't had time to completely explore.}}
\item authors in~\cite{vekiri2002value} is lit review discussing the different theories about the utility of graphics in learning (dual coding theory, visual argument hypothesis, and conjoint retention hypothesis). The article contains a nice definition of graphics (though it might be too domain specific for statistical graphics, since it claims graphics have only one meaning and aren't prone to interpretation.) Depending on the context, ``maps", ``charts", and ``graphs" here could fall into a loose classification as ``statistical graphics", but in general ``graphs" is probably accurate for most of the lineups we're testing: ``[Referents are ] quantitative data ... that enable viewers to compare and observe relations among variables". 
